### 2023-03-20 paper-thesis/chapters/23-adding-universes.tex
Update 23 and 24

Our language so far has the property that the type of \verb|Type| is \verb|Type|, this is called \emph{type in type}. This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent: we can prove false.

@@

The solution to this problem is introducing universes into the language. We now have the following hierarchy of types:
\begin{lstlisting}
true : Bool : Type 0 : Type 1 : Type 2 : ...
\end{lstlisting}

@@

\section{Implementing Universes in the CoC}

The constructor of Type has been changed to \verb|Type : int -> Expr|. We need to update all occurrences of \verb|Type| in figure \ref{fig:type-check-rules}. Figure \ref{fig:type-check-rules-universes} contains all the rules that require a change.

The three rules that were changed are:
\begin{enumerate}
	\item The rule for type-checking a \verb|Type| expression, which now is explicit about the universes of each \verb|Type|.
	\item The rule for type-checking \verb|FnType|, which takes the maximum of the two arguments.
	\item The rule for type-checking \verb|FnConstruct|, which ignores the universe of its argument type, and only verifies that it is a type.
\end{enumerate}

@@

		% FnType rule
		\inferrule{
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}(u_a)}
			\\ \bred{s}{a}{a'}
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}(u_b)}
		} {
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}(max(u_a, u_b))}
		}

@@

>>>>

Our language so far has the property that the type of \verb|Type| is \verb|Type|, this is called \emph{type in type}. This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent: we can prove false~\cite{girard}.

@@

The solution to this problem is to introduce universes into the language. We now have the following hierarchy of types:
\begin{lstlisting}
true : Bool : Type 0 : Type 1 : Type 2 : ...
\end{lstlisting}

@@

\section{Implementing Universes in the Calculus of Constructions}

The \verb|Type| constructor has to be changed to \verb|Type : int -> Expr|, taking the universe of the type as an argument. We then need to update all occurrences of \verb|Type| in figure \ref{fig:type-check-rules}. Figure \ref{fig:type-check-rules-universes} contains all the rules that require a change.

The three rules that were changed are:
\begin{enumerate}
	\item The rule for type-checking a \verb|Type| expression. The universe of \verb|Type u| is \verb|u + 1|. 
	\item The rule for type-checking \verb|FnType|. The universe of \verb|FnType| is the maximum of the universe of the argument and return type.
	\item A trivial change to the rule for type-checking \verb|FnConstruct|, which ignores the universe of its argument type.
\end{enumerate}

@@

		% FnType rule
		\inferrule{
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}(u_a)}
			\\ \bred{s}{a}{a'}
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}(u_b)}
		} {
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}(\co{max}(u_a, u_b))}
		}

@@

We have now finished the type checker of the language. Next, we will discuss how to implement the back end of the language.

### 2023-03-20 paper-thesis/chapters/24-backend.tex
Update 23 and 24

\section{Interpreter}
It is valuable to define an interpreter for our language, because it is a way to run a program without relying on the definition of the language you are compiling to. For a dependently typed language specifically it is easy to add on, as we already needed to define the dynamic semantics of our language for type checking anyways.

@@

An advantage of implementing the dynamic semantics in Statix instead of Stratego is that we can use scope graphs as a part of our dynamic semantics rules.

A disadvantage is that Stratego is specifically optimized for these transformations~\cite{fcpmc}. Therefore, the resulting interpreter will be slower when implemented in Statix. \todo{comparison}

\section{Compiler}
The alternative to writing an interpreter is writing a compiler to another language. In order to explore this avenue, we wrote a compiler to Clojure, a dynamically typed dialect of Lisp that runs on the JVM~\cite{clojure}. The advantage of a compiling to such a high-level language is that the actual compilation steps are relatively straight-forward, but it still shows that we can write a compiler for a dependently typed language in Spoofax.

Both languages support first-class functions, so we can map most constructs of our language directly onto constructs of Clojure. Therefore, most of our compiler has rules that look like the example below, where the constructors are compiled recursively using the \verb|compile-expr| strategy, ignoring the specified types. The constructors starting with \verb|C| are constructors from Clojure. 
\begin{lstlisting}
compile-expr : FnConstruct(x, _, b) -> 
  CFn(<compile-id> x, <compile-expr> b)
\end{lstlisting}

Types require a bit more attention. Since our language has no way to match on types, types can actually be completely eliminated during type checking. The easiest way to do this is to compile all types to an arbitrary value (we chose the integer 0), since the value is irrelevant anyways. A more efficient solution could be to implement type erasure~\cite[Section 23.6]{tapl}, where type-level functions are completely removed. We decided against this as the goal of this section is to show that we can compile the language with ease, not to write an efficient compiler.

-analysis is hard, agda doesn't do it
- if compiled language doesn't matter
- erasure annotations (idris/agda)

Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
- agda to scheme compiler
- atoms voor constructors

>>>>

\section{Interpreter}
It is valuable to define an interpreter for our language, because it is a way to run a program without relying on another language to compile to. For a dependently typed language specifically, it is easy to develop an interpreter, as we already needed to define the dynamic semantics of our language for type checking anyways.

@@

We defined the dynamic semantics in Statix, which comes with some advantages and disadvantages over defining it in Stratego.

An advantage of implementing the dynamic semantics in Statix instead of Stratego is that we can use scope graphs as a part of our dynamic semantics rules. As discussed in chapter~\ref{chap:baselang}, scope graphs are a natural way to store the environment of the language during evaluation. When dynamic semantics is defined in Stratego, a different representation of the environment would be needed.

A disadvantage is that Stratego is a domain specific language for transformations, and it has some features that would make the definitions of beta reductions shorter. For example, to define \verb|betaReduce| in terms of \verb|betaReduceHead| we need to match the constructor and apply \verb|betaReduceHead| recursively. Stratego has the \verb|topdown| traversal strategy that does the same thing without needing to be explicitly defined on all constructors.

Another disadvantage is that Stratego is specifically optimized for these transformations, for example as discussed in "Optimising First-Class Pattern Match Compilation"~\cite{fcpmc}. Therefore, the resulting interpreter will be slower when implemented in Statix. This is however just a constant factor overhead, it does not affect the time complexity of the algorithm.

\section{Compiler}
An alternative to writing an interpreter is writing a compiler to another language. In order to explore this avenue, we wrote a compiler to Clojure, a dynamically typed dialect of Lisp that runs on the JVM~\cite{clojure}. The advantage of a compiling to such a high-level language is that the actual compilation steps are relatively straight-forward, but it still shows that we can write a compiler for a dependently typed language in Spoofax.

The compiler is created while keeping in mind the Agda to Scheme compiler, available at \url{https://github.com/jespercockx/agda2scheme}. This is also a compiler from a dependently typed language (Agda) to a LISP (Scheme). We make some of the same choices as this compiler, as these choices have been proven to work.

\subsection{Calculus of Constructions}
Both languages support first-class functions, so we can map most constructs of our language directly onto constructs of Clojure. Therefore, most of our compiler rules look like the example below, where the constructors are compiled recursively using the \verb|compile-expr| strategy, ignoring the specified types. The constructors starting with \verb|C| are constructors from Clojure. 
\begin{lstlisting}
compile-expr : FnConstruct(x, _, b) -> 
  CFn(<compile-id> x, <compile-expr> b)
\end{lstlisting}

Types require a bit more attention. Since our language has no way to match on types, types can actually be completely eliminated during type checking. The easiest way to do this is to compile all types to an arbitrary value (we chose the string "TYPE"), since the value is irrelevant anyways. A more efficient solution could be to implement type erasure~\cite[Section 23.6]{tapl}, where type-level functions are completely removed. We decided against this as the goal of this section is to show that we can compile the language with ease, not to write an efficient compiler. This optimization was also not implemented in the Agda to Scheme compiler.

\subsection{Inductive Data Types}

Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
- atoms voor constructors
